Motional Quantum Ground-State Cooling Outside the Lamb-Dicke Regime

### 2017-06-25 paper.tex
Change how optical pumping heating is explained

\begin{figure*}
  \includegraphics[height=5.2cm]{imgs/Na_RSC_schematic.pdf}
  \includegraphics[height=4.5cm]{sequence.pdf}
  \caption{(A) Energy levels and schematic of Raman sideband cooling.
    The Raman transitions have a one photon detuning $\Delta=25$ GHz from the D2 line.
    We use D1 light with $\sigma^-$ polarization to repump atoms out of $|F=2,m_F=-1\rangle$
    state to minimize heating on the atom in $|F=2,m_F=-2\rangle$ state
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.
    (C) Schematic of the cooling sequence. The tweezer switches at 3 MHz to
    reduce light shifts during optical pumping. Each cooling cycle consists of $8$ pulses.
    The four axial pulses are addressing two neighboring cooling orders.
    The two pulses in each radial directions are either addressing two neighboring cooling orders
    or having different length on the first order when most of the population are below $n=3$
    towards then end of the cooling sequence.
    \label{f-setup}}
\end{figure*}
\begin{figure}[b]
  \includegraphics[width=8.5cm]{imgs/fig2_raman_op.pdf}
  \caption{Matrix elements and heating probabilities as a function of motional state.
    The range plotted covers $99\%$ of the initial thermal distribution.
    (A) Matrix elements for Raman transition in the axial direction showing deviation from
    $\sqrt{n}$ scaling and multiple minimums for different sideband orders.
    (B) Heating probability during the optical pumping step for all three axis.
    Due to the large Lamb-Dicke parameter,
    there is a high probability of heating especially in the axial direction.
    \label{f-ld}}
\end{figure}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_r.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_p1.pdf}
  \caption{(A) Radial Raman sideband spectrum of first order heating, first order cooling and
    second order cooling before and after Raman sideband cooling.
    (B,C) Rabi flopping on axis 3 (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue lines corresponds to a ground state probability of $93\%$ after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    \label{f-radial}}
\end{figure*}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_a1.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_p1.pdf}
  \caption{(A) Axial Raman sideband spectrum from first order heating to eighth order cooling
    before and after Raman sideband cooling.
    The data for the second and higher orders of cooling sidebands are taken with 150 $\mu$s
    pulse time and the rest are taken with 125 $\mu$s pulse time.
    (B,C) Rabi flopping on axial (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue and green lines corresponds to a ground state probability of $92\%$ after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    The blue lines does not take into account the effect of decoherence due to resonance
    fluctuation. By comparing it to the green line in (C), which includes a $3$ kHz fluctuation,
    we can clearly see that this effect is the strongest for the after cooling data on
    the axial heating sideband where the Rabi frequency is the lowest.
    \label{f-axial}}
\end{figure*}

@@

In order to achieve long coherence times and full quantum control of the system,
it is typically necessary to cool the atom to the
three dimensional motional ground state in the optical tweezer.
This has been demonstrated for neutral Rb\cite{Thompson2013,Kaufman2012} and Cs\cite{Liu2017}
in experiments that had low initial temperature and small Lamb-Dicke parameter.
These features may not be easily achievable for other systems,
such as directly laser-cooled polar molecules \fxnote{REF},
or atoms with challenging sub-Doppler cooling such as Na, K, and Li \fxnote{REF}.
In this letter, we cool single sodium atoms trapped in optical tweezers to
the motional ground state with Raman sideband cooling.
Despite having a large Lamb-Dicke parameter and high initial temperature,
we are able to achieve a ground state probability of $77(4)\%$
by utilizing several new cooling techniques and a carefully optimized cooling sequence.
Our approach is quite general, and opens up ground-state cooling for other systems.

@@

In order to perform Raman sideband cooling efficiently and minimize
the heating during the optical pumping, we need to have a low initial temperature and
a small LD parameter. However, since the LD parameter $\eta\propto 1/(\sqrt[4]{m}\lambda)$ where $m$ is the mass of the atom and $\lambda$
is the wavelength of the atomic transition,
it is larger for sodium at a given trap depth due to its low mass and short D line wavelength.
With 45 mW of power in the trap, we measure trapping frequencies of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69(1), 430(4), 590(5)\}\ \text{\text{kHz}}$
which corresponds to optical pumping LD parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.602(5), 0.241(2), 0.206(1)\}$
Moreover, due to the small hyperfine splitting in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of 70 $\mu$K. Combined with the high $\eta$, this gives us a initial effective optical pumping LD parameters of
$\{\eta^{OP}_{1\eff},\eta^{OP}_{2\eff},\eta^{OP}_{3\eff}\} = \{4.0(1), 0.67(2), 0.50(1)\}$
As a result, there is a very high probability of heating during the optical pumping step
(figure \ref{f-ld}B) causing an estimated $30\%$ average heating probability during optical pumping
in the weakly confined axial direction \fxnote{So what?  Is 30\% too high?  Can we interpret this in terms of an observable, like best-possible ground state fraction?}.

>>>>

\begin{figure*}
  \includegraphics[height=5.2cm]{imgs/Na_RSC_schematic.pdf}
  \includegraphics[height=4.5cm]{sequence.pdf}
  \caption{(A) Energy levels and schematic of Raman sideband cooling.
    The Raman transitions have a one photon detuning $\Delta=25$ GHz from the D2 line.
    We use D1 light with $\sigma^-$ polarization to repump atoms out of $|F=2,m_F=-1\rangle$
    state to minimize heating on the atom in $|F=2,m_F=-2\rangle$ state
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.
    (C) Schematic of the cooling sequence. The tweezer switches at 3 MHz to
    reduce light shifts during optical pumping. Each cooling cycle consists of $8$ pulses.
    The four axial pulses are addressing two neighboring cooling orders.
    The two pulses in each radial directions are either addressing two neighboring cooling orders
    or having different length on the first order when most of the population are below $n=3$
    towards then end of the cooling sequence.
    \label{f-setup}}
\end{figure*}
\begin{figure}[b]
  \includegraphics[width=8.5cm]{imgs/fig2_raman_op.pdf}
  \caption{Matrix elements and heating as a function of motional state.
    The range plotted covers $95\%$ of the initial thermal distribution.
    (A) Matrix elements for Raman transition in the axial direction showing deviation from
    $\sqrt{n}$ scaling and multiple minimums for different sideband orders.
    (B) Average change in motional state after the optical pumping step for all three axis.
    Due to the large Lamb-Dicke parameter,
    there is a high probability of $n$ changing especially in the axial direction.
    \label{f-ld}}
\end{figure}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_r.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_p1.pdf}
  \caption{(A) Radial Raman sideband spectrum of first order heating, first order cooling and
    second order cooling before and after Raman sideband cooling.
    (B,C) Rabi flopping on axis 3 (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue lines corresponds to a ground state probability of $93\%$ after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    \label{f-radial}}
\end{figure*}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_a1.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_p1.pdf}
  \caption{(A) Axial Raman sideband spectrum from first order heating to eighth order cooling
    before and after Raman sideband cooling.
    The data for the second and higher orders of cooling sidebands are taken with 150 $\mu$s
    pulse time and the rest are taken with 125 $\mu$s pulse time.
    (B,C) Rabi flopping on axial (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue and green lines corresponds to a ground state probability of $92\%$ after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    The blue lines does not take into account the effect of decoherence due to resonance
    fluctuation. By comparing it to the green line in (C), which includes a $3$ kHz fluctuation,
    we can clearly see that this effect is the strongest for the after cooling data on
    the axial heating sideband where the Rabi frequency is the lowest.
    \label{f-axial}}
\end{figure*}

@@

In order to achieve long coherence times and full quantum control of the system,
it is typically necessary to cool the atom to the
three dimensional motional ground state in the optical tweezer.
This has been demonstrated for neutral Rb\cite{Thompson2013,Kaufman2012}
and Cs\cite{Li2012,Liu2017}
in experiments that had low initial temperature and small Lamb-Dicke parameter.
These features may not be easily achievable for other systems,
such as directly laser-cooled polar molecules \fxnote{REF},
or atoms with challenging sub-Doppler cooling such as Na, K, and Li \fxnote{REF}.
In this letter, we cool single sodium atoms trapped in optical tweezers to
the motional ground state with Raman sideband cooling.
Despite having a large Lamb-Dicke parameter and high initial temperature,
we are able to achieve a ground state probability of $77(4)\%$
by utilizing several new cooling techniques and a carefully optimized cooling sequence.
Our approach is quite general, and opens up ground-state cooling for other systems.

@@

In order to perform Raman sideband cooling efficiently and minimize
the heating during the optical pumping, we need to have a low initial temperature and
a small LD parameter. However, since the LD parameter $\eta\propto 1/(\sqrt[4]{m}\lambda)$ where $m$ is the mass of the atom and $\lambda$
is the wavelength of the atomic transition,
it is larger for sodium at a given trap depth due to its low mass and short D line wavelength.
With 45 mW of power in the trap, we measure trapping frequencies of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69(1), 430(4), 590(5)\}\ \text{\text{kHz}}$
which corresponds to optical pumping LD parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.602(5), 0.241(2), 0.206(1)\}$
Moreover, due to the small hyperfine splitting in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of 70 $\mu$K. Combined with the high $\eta$, this gives us a initial effective optical pumping LD parameters of
$\{\eta^{OP}_{1\eff},\eta^{OP}_{2\eff},\eta^{OP}_{3\eff}\} = \{4.0(1), 0.67(2), 0.50(1)\}$
As a result, there is a very high average change of motional state during the optical pumping step
(figure \ref{f-ld}B). The averaged change in motional states over the initial distribution
is $2.4$ in the axial direction.
Compare to $0.89-0.95$ in the axis with the weakest confinement in previous experiments
\cite{Thompson2013,Kaufman2012,Li2012,Liu2017} this causes greater uncertainty in the motional
state making it harder to cool.
